Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-03-20 paper/finalish/article_1023.tex
updates to include attribution pre/post intervention

The source attribution study described here uses human case data ($H=1,400$) collected in the Manawatu region of New Zealand from March 2005 to December 2014. In addition, over the same time period and from the same geographical location, isolates $N=2,080$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others. The total number of unique genotypes from all isolates is $I=355$, with 35\% of genotypes found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases. As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants\cite{Muell, Coll, Cart}. The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.

The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random. Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census\cite{Stat1, Stat2}. The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures/attr_all.pdf}
\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig2}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures/genotype_figure.pdf}
\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
\label{fig:genotype_dist}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures/attr_intervention.pdf}
\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area. The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig:intervention}
\end{figure}

@@

As noted previously, a major public health initiative in 2007 led to a significant reduction in the number of cases of campylobacteriosis in New Zealand. In order to examine the effects of this change on attribution probabilities, we repeated the analysis by including an interaction with time period (2005--2007 and 2008--2014). Figure~\ref{fig:intervention} shows that the general trend in attribution by rurality for each of the time periods utilising a linear trend on the logit scale to incorporate rurality.

TODO: FILL THIS IN...

is consistent with that seen with all the data, with ruminant associated campylobacteriosis in highly rural areas to poultry associated cases in urban centers linked by a linear trend. However, the proportion attributed to poultry across all rurality levels decreased (and thus attribution to ruminants increased) in this post-intervention dataset, being most marked in urban centers. Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.

>>>>

The source attribution study described here uses human case data ($H=1,400$) collected in the Manawatu region of New Zealand from March 2005 to December 2014. In addition, over the same time period and from the same geographical location, isolates $N=2,080$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others. The total number of unique genotypes from all isolates is $I=355$, with 35\% of genotypes found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases. As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants \cite{Muell, Coll, Cart}. The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.

The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random. Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census \cite{Stat1, Stat2}. The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/attr_all.pdf}
\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig2}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/genotype_figure.pdf}
\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
\label{fig:genotype_dist}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/attr_intervention.pdf}
\caption{Posterior mean attribution ($F$) of human cases from 2005--2007 and 2008--2014 with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area. The attribution is generated using the linear model given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig:intervention}
\end{figure}

@@

As noted previously, a major public health initiative in 2007 led to a significant reduction in the number of cases of campylobacteriosis in New Zealand. In order to examine the effects of this change on attribution probabilities, we repeated the analysis by including an interaction with time period (2005--2007 and 2008--2014). Figure~\ref{fig:intervention} shows that the general trend in attribution by rurality for each of the time periods utilising a linear trend on the logit scale to incorporate rurality. There is a clear difference, with a significantly lower attribution to poultry (and correspondingly higher attribution to ruminants) in all but the most rural of areas, being strongest in highly urban areas. Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
